An Oral History Project: Coke Studio and the Revitalization of Eastern Music

### 2022-05-09 paper.tex
just final proofread and I'm done

\title{An Oral History Project: Coke Studio and the Revitalization of Eastern Music}
\date{April 2022}
\author{Sarim Aleem}

@@

  Like Rohail Hyatt, most Pakistani's aren't aware of the one thousand year
  heritage of music and culture that in the diaspora. Most would ignore the rich
  classical music scene for being boring or inaccessible, and would most likely
  rather opt for bollywood style dance videos, a dime a dozen in the country.

  But as Hyatt, the leader of a reignition of a rich and vibrant traditional
  culture would discover, there is an incredible philosophical and poetic depth
  to traditional music.  But little did he expect that the reignition of that
  lost culture would not come from the local film or entertainment industry.
  Rather, it would come from a well known Western soft drink company, initially
  trying to outcompete its famed rival through a small marketing stint: Coca
  Cola. And, he would spearhead it.

  The introduction of Coke Studio in Pakistan has led to the revitalization of
  traditional culture by putting a reintroducing endangered languages, putting a
  spotlight on traditional Eastern instruments, and bringing traditional Sufi
  culture into the modern day.

@@

  Thus, Coke studio was born; blending East and West when it came to musical
  instruments, culture, and musical character. This blend would not only
  revitalize traditional culture but innovate on it.

@@

  There are three main branches in the religion of Islam. Sunni - the normative
  version of Islam practiced by the majority of the world. Shi'a - a smaller but
  not insignificant sect with minor differences in religious interpretation. And
  finally, there is Sufi Islam. Although technically Sufi Islam is not
  considered by some to be a separate branch of Islam, many of their practices
  are not found in normative Islam and thus, for the purposes of this paper they
  will be considered a different "branch" or "sect".

  Sufi Islam is commonly defined "Islamic Mysticism" and tends to emphasize the 
  spiritual aspects of the religion more so than the legalistic, or worldy
  elemnts. For most orientalists, the religion is thus seen as the more 
  moderate, "peaceful" interpretation of islam \autocite{beg2020fizzy}. This of
  course incorrectly implies that traditional mainstream islam is oppressive,
  but nevertheless the perception of Westerners ironically have a large
  influence on what local Pakistanis wish to express their culture. As a
  consequence, embracing the Sufi genre of music - the Qawwali - is a nice
  middle ground in appeasing to a moderate image of Islam while simultaneously 
  appealing to the more religiously inclined in the nation who are skeptical of
  the value of music in society \autocite{beg2020fizzy}.

@@

@@

  Part of the reason that the genre of Sufi music became so popular was that it
  largely touched on the spiritual aspects of religion. Another part is that the
  poem predates the beginning of political differences in the Indian
  subcontinent. And yet another is the fact that The video is freely available
  on youtube, arguably one of the largest if not the largest video streaming
  site in the world. Combined with Coca Cola's large marketing budget, it's not then 
  surprising the song became the hit it did.

@@

  Afreen is a quintissential example of Coke studio subverting expectations when
  it came to artist selection. The song was originally sung by the dubbed "King
  of Qawwali" Nusrat Fateh Ali Khan. The cover on the other hand was not only
  sung by his nephew, the esteemed Rahat Fateh Ali Khan, but also by Momena
  Mostehsan: a young woman raised in the United States. The combining of
  cultures brings Sufi culture into the modern era, and thus there's a sense of
  innovation in the music. Furthermore, not only is the artist selection crucial,
  but also their appearance.

\begin{quote}
  what garnered more attention sometimes were the looks of the artist, than the
  artist himself/ herself. One would often find the folk musician accompanied by a
  female singer in jeans, cropped leather jacket, painted nails, which was indeed
  a counter narrative to the burkha wearing Islam women (\autocite{Chibmedium}).
\end{quote}

Thus, although Coke Studio brings old Sufi culture into the modern day, it is not done
without changing who performs the songs.

\section{Homogenization of Language}

@@

Unlike the United States, which is mostly homogenous in terms of linguistic
diversity, the hindustani region has an incredible amount of languages and
subcultures splintered in.  There are over 30 languages alone with over 1
million speakers in the subcontinent, and about 122 languages with more than
10,000 speakers.

@@

  Fifteen languages split the remaining spots. In comparison to Punjabi, the other
  three major regional languages pale: while Pashto, Balochi and Sindhi together
  account for over one third of native speakers in Pakistan, they only amount to
  four percent of the linguistic presence in Coke Studio Pakistan.
\end{quote}

It's hard to forget that at the end of the day, Coca Cola is a business. It's
simply more convenient to appeal to the "prestige" languages that are used in
Poetry (Persian) or Pop Culture (Urdu, Punjabi) than go out of the way to produce 
music in other languages.

@@

 It's not just that certain nuances are getting lost in what these artists are
 singing. They are actually singing some completely meaningless lines and you
 can imagine what that means for the transmission of meaning for classical
 poetry that has been traditionally passed on from one generation to another.

@@

Although the Coke Studio discussed, the one popularly known, is the Pakistani
one, Coca Cola has in fact attempted to establish studios in other countries,
including India. However, the Indian Coke studio never reached the fame the
Pakistani did, since it couldn't escape it's consumerist, capitalist, Bollywood
roots. But, the success of Pakistani Coke Studio points to the fact that in the
face of the globalization and westernization of Pakistan, there is a sense of
pride in sublime of classic poetry and music, however imperfect the modern
rendition of that may be.

>>>>

\title{An Oral History Project: Coke Studio and the Revitalization of Eastern Music}
\date{April 2022}
\author{Sarim Aleem \\ Word Count: 2135}

@@

  Like Rohail Hyatt, most Pakistani's aren't aware of the one thousand year
  heritage of music and culture that in the diaspora. Most would ignore the rich
  classical music scene for being boring or inaccessible, and would most likely
  rather opt for bollywood style dance videos - a dime a dozen in the country.

  But as Hyatt, the leader of a reignition of a rich and vibrant traditional
  culture would discover, there is an incredible philosophical and poetic depth
  to traditional music.  But little did he expect that the reignition of that
  lost culture would not come from the local film or entertainment industry.
  Rather, it would come from a well known Western soft drink company, initially
  trying to outcompete its famed rival through a small marketing stint: Coca
  Cola.

  The introduction of Coke Studio in Pakistan has led to a revitalization of
  classical music by reintroducing Sufi culture, putting a diverse catalog of
  artists in music, and having a nontrivial amount of linguistic diversity in
  songs.  Nevertheless, the production studio has it's flaws characteristic of a
  large capitalistic coorporation, and oftentimes commoditizes classic culture
  and neglects it's heritage.

@@

  And then, Coke studio was born; blending East and West when it came to musical
  instruments, culture, and musical character. This blend would not only
  revitalize traditional culture but for better of worse, innovate on it.

@@

  There are three main branches in the religion of Islam. Sunni - the normative
  version of Islam practiced by the majority of the world. Shi'a - a smaller but
  not insignificant sect with minor differences in religious interpretation. And
  finally, there is Sufi Islam. Although technically Sufi Islam is not
  considered by some to be a separate branch of Islam, many of their practices
  are not found in normative Islam and thus, for the purposes of this paper they
  will be considered a different "branch" or "sect" \autocite{sufibackground}.

  Sufi Islam is commonly defined "Islamic Mysticism" and tends to emphasize the 
  spiritual aspects of the religion more so than the legalistic, or worldy
  elemnts. For most orientalists, the religion is thus seen as the more 
  moderate, "peaceful" interpretation of islam. This of
  course incorrectly implies that traditional mainstream islam is oppressive,
  but nevertheless the perception of Westerners ironically have a large
  influence on what local Pakistanis wish to express their culture \autocite{beg2020fizzy}. As a
  consequence, embracing the Sufi genre of music - the Qawwali - is a nice
  middle ground in appeasing to a moderate image of Islam while simultaneously 
  appealing to the more religiously inclined in the nation who are skeptical of
  the value of music in society \autocite{williams2019soundtrack}.

@@

  As with most Sufi poetry, the lyrical content of Chaap Tilak is very divorced from 
  political reality, and is far more focused on abstract and "higher" concepts:

    The Game of Love
    The Game of Love
    If I play with my dear one

    If I win, My sweetheart is Mine
    If I lose, I'm still with my Dear one

    \dots

    The night of the Prophet's celebration has dawned
    The night has dawned
    The night has dawned

    \dots

    With one glance your enchanting eyes

  Love and devotion to God featured heavily in the song, and much of Sufi
  Qawwali is described as eschewing wordly and material life - hardly a
  controversial message. For this reason, Qawwali's are a popular genre to
  market to worldwide audiences

@@

  That's not to say that everyone is happy with the reintroduction of Sufism in modern culture. 
  For many, the reintroduction of Sufi culture seems commodotized. As Amin writes:

  \begin{quote}
    What happened to qawwali in this process? [After it's greater presence in
    90's] It was unmoored from its deeper spiritual significance and repackaged
    as a cultural item, all so that it could be more easily commodified for a
    global audience. \autocite{sellingsufism}
  \end{quote}

  For many, Coke Studio adopting the Qawwali was not seen as a victory, but
  rather a capitalist society appropriating a millenia old tradition for profit.
  In fact, although the Qawwali comes from the Islamic tradition, "Qawwali is
  rarely marketed as Islamic music because to do so would risk turning away
  consumers, as Islam purportedly does not share the same peaceful values as
  Sufism" \autocite{sellingsufism}. So in essence, a company like Coke Studio 
  takes the Qawwali, and erases it's cultural heritage when convenient to make 
  profits - Hardly an ode to authenticity.

  Part of the reason that the genre of Sufi music became so popular was that it
  largely touched on the spiritual aspects of religion. Another part is that the
  poem predates the beginning of political differences in the Indian
  subcontinent. And yet another is the fact that The video is freely available
  on youtube, arguably one of the largest if not the largest video streaming
  site in the world. Combined with Coca Cola's large marketing budget, it's not then 
  surprising that Qawwalis became the hits they did in Coke Studio.

@@

  Afreen is a quintissential example of Coke studio subverting expectations when
  it came to artist selection. The song was originally sung by the dubbed "King
  of Qawwali" Nusrat Fateh Ali Khan. The cover on the other hand was not only
  sung by his nephew, the esteemed Rahat Fateh Ali Khan, but also by Momena
  Mostehsan: a young woman raised in the United States \autocite{afreen}. The combining of
  cultures brings Sufi culture into the modern era, and thus there's a sense of
  innovation in the music. Furthermore, not only is the artist selection crucial,
  but also their appearance.

\begin{quote}
  what garnered more attention sometimes were the looks of the artist, than the
  artist himself/ herself. One would often find the folk musician accompanied by a
  female singer in jeans, cropped leather jacket, painted nails, which was indeed
  a counter narrative to the burkha wearing Islam women \autocite{Chibmedium}.
\end{quote}

The inclusion of women and nontraditional artists is important because it
signals a change in how classical music is performed. The music is stripped down
to it's bare minimum and reconstructed. And, for the most part it has succeeded.
Songs with these new and modern artists have hundreds of millions of views on
youtube, and have captivated the imagination of the diaspora.

\section{The language of Coke Studio}

@@

Unlike the United States, which is mostly homogenous in terms of linguistic
diversity, the Hindustani region has an incredible amount of languages and
subcultures splintered in.  There are over 30 languages alone with over 1
million speakers in the subcontinent, and about 122 languages with more than
10,000 speakers.

@@

  Fifteen languages split the remaining spots. In comparison to Punjabi, the other
  three major regional languages pale: while Pashto, Balochi and Sindhi together
  account for over one third of native speakers in Pakistan, they only amount to
  four percent of the linguistic presence in Coke Studio Pakistan \autocite{soundscape}.
\end{quote}

That's not to say that Coke Studio has no linguistic diversity. The producers
are humans after all and do make attempts to put in some sort of linguistic
diversity.  For example the following languages and scripts have been learnt and
used by the producers: "Seraiki, Sindhi, Pashto, Punjabi, Brahui, Balochi, Braj
Bhasha, Marwari, Bengali, Urdu, Persian, Arabic and Turkish" \autocite{herald}.
That is still a great deal more than is present in bollywood media.

Still, It's hard to forget that at the end of the day, Coca Cola is a business. It's
simply more convenient to appeal to the "prestige" languages that are used in
Poetry (Persian) or Pop Culture (Urdu, Punjabi) than go out of the way to produce 
music in other languages \autocite{ethnographicInterview}.

@@

\begin{quote}
 It's not just that certain nuances are getting lost in what these artists are
 singing. They are actually singing some completely meaningless lines and you
 can imagine what that means for the transmission of meaning for classical
 poetry that has been traditionally passed on from one generation to another \autocite{herald}.
\end{quote}

We tend to see unfortunately that there is a sense of neglect. One megahit that 
coke studio produced is a song by the name "Taj  Da Re Haram"
\autocite{taajdaar}. The song is the most popular song on the youtube channel by
view count (over 374 million views as of this writing).  However, out of
carelessness many of the lyrics are wrong to the point of being nonsensical
\autocite{herald}.

In some part, this failing can be attributed to the fact that education in
traditional languages like Urdu have lessened in recent years, as English is a
far more useful language from a capitalistic standpoint. Popular artists don't 
fact check their songs for accuracy all the time. This is reminiscent of the 
previous discussion on the commodotization of Qawwali's. After all, if a great 
deal of the population is unable to fact check the songs themselves, then it
doesn't hurt the bottomline too much to overlook careful scrutiny that ensures
the classic renditions are authentic.

@@

Although the Coke Studio discussed and the one popularly known is the Pakistani
one, Coca Cola has in fact attempted to establish studios in other countries,
including India. However, the Indian Coke studio never reached the fame the
Pakistani did, since it couldn't escape it's consumerist, capitalist, Bollywood
roots. But, the success of Pakistani Coke Studio points to the fact that in the
face of the globalization and westernization of Pakistan, there is a sense of
pride in sublime of classic poetry and music, however imperfect the modern
rendition of that may be.

Coke Studios innovation in the field has been nothing short of incredible.
Classic Poetry from almost a millenia ago are being popularized and being
listened to hundreds of millions of times.  However, in the usage of classical
sources, it is imperative to stay faithful to them, which Coke Studio has failed
to do. After all, for many, Coke Studio has become more than a marketing stint; 
it has trancended to Pakistan's oral history project.
